---
title: "S&DS363_PSET_2"
author: "Ethan Mathieu, Tony Ruan, Humrya Karim"
date: "2024-02-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("biotools")
install.packages("klaR")
install.packages("ggExtra")
install.packages("heplots")
install.packages("rgl")

library(MASS)
library(biotools)
library(klaR)
library(car)
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggExtra)
library(heplots, exclude = "rgl")

```



``` {r data_set_up}
GED <- read.csv("/Users/Tony/Desktop/class docs/semester 6/S&DS 363/Global_Education.csv")

GED_Leveled <- GED |>
   mutate(ed_level = case_when(
    ((Completion_Rate_Upper_Secondary_Male + Completion_Rate_Upper_Secondary_Female) / 2) < 50 ~ "1",
    ((Completion_Rate_Upper_Secondary_Male + Completion_Rate_Upper_Secondary_Female) / 2) >= 50 &
    ((Completion_Rate_Upper_Secondary_Male + Completion_Rate_Upper_Secondary_Female) / 2) < 80 ~ "2",
    ((Completion_Rate_Upper_Secondary_Male + Completion_Rate_Upper_Secondary_Female) / 2) >= 80 ~ "3"
  )) |>
    select(Completion_Rate_Upper_Secondary_Female, Completion_Rate_Upper_Secondary_Male, Birth_Rate, Unemployment_Rate, ed_level)
```

### Question 1

```{r question_1}

par(mfrow = c(1,2), pty = "s", cex = 0.8)
boxplot(GED_Leveled$Birth_Rate ~ GED_Leveled$ed_level)
boxplot(GED_Leveled$Completion_Rate_Upper_Secondary_Female ~ GED_Leveled$ed_level)
boxplot(GED_Leveled$Completion_Rate_Upper_Secondary_Male ~ GED_Leveled$ed_level)
boxplot(GED_Leveled$Unemployment_Rate ~ GED_Leveled$ed_level)
par(mfrow = c(1, 1))

par(mfrow = c(1,3), pty = "s", cex = 0.8)
cqplot(GED_Leveled[GED_Leveled$ed_level == "1", c("Birth_Rate","Completion_Rate_Upper_Secondary_Female", "Completion_Rate_Upper_Secondary_Male", "Unemployment_Rate")], main = "Low")
cqplot(GED_Leveled[GED_Leveled$ed_level == "2", c("Birth_Rate","Completion_Rate_Upper_Secondary_Female", "Completion_Rate_Upper_Secondary_Male", "Unemployment_Rate")], main = "Medium")
cqplot(GED_Leveled[GED_Leveled$ed_level == "3", c("Birth_Rate","Completion_Rate_Upper_Secondary_Female", "Completion_Rate_Upper_Secondary_Male", "Unemployment_Rate")], main = "High")
par(mfrow = c(1, 1))

boxM(GED_Leveled[,c("Birth_Rate","Completion_Rate_Upper_Secondary_Female", "Completion_Rate_Upper_Secondary_Male", "Unemployment_Rate")], GED_Leveled$ed_level)

plot(GED_Leveled$Birth_Rate, GED_Leveled$Completion_Rate_Upper_Secondary_Female, col = as.numeric(GED_Leveled$ed_level)+1, pch = 19, main = "World Educational Relationships", ylab = "Female Completion Rate", xlab = "Birth Rate")

plot(GED_Leveled[,1:4], col = GED_Leveled$ed_level, pch = 19, cex = 1.2)
```
* We examined 4 variables in relation to our global education data levels: Birth Rate, Unemployment Rate, Male educational completion, and Female educational completion. Within each category of discrimination, which we determined to be high, medium, and low rates of education, each group appears to have a multivariate normal distribution according to our chi-squared quantile plots. It appears that for countries with different levels of educational attainment broadly have different means from one another for the level of female school completion and male school completion—while unemployment rate and birth rate have more similar means. Examining the scatterplots, as well as performing Box's M-test—which yielded a very low p-value, there definitely appears to be different means between the three levels of low-medium-high However, buy also  different covariances between the three groups. As a result, we intend to utilize quadratic discrimiination analysis to account for these different variances. As a side note, we quickly would like to add that we observe different numbers of observations between each level of literacy—with more countries being classified as low-literacy rates than medium/high. We will account for this in our classification analysis.

### Question 2

```{r question_2}

# Linear Discrimnant Analysis
(GED_Leveled_disc_lda <- lda(GED_Leveled[,1:4], grouping = GED_Leveled$ed_level))
(GED_Leveled_disc_qda <- qda(GED_Leveled[,1:4], grouping = GED_Leveled$ed_level))

#Wilks Significance Test
GEDtest <- qda(GED_Leveled[, -5], grouping = GED_Leveled$ed_level)
summary(GEDtest)
GED.manova <- manova(as.matrix(GED_Leveled[, -5]) ~ GED_Leveled$ed_level)
summary.manova(GED.manova, test = "Wilks")
summary.aov(GED.manova)
```
* We ran both linear discriminant analysis and quadratic discriminant analysis and cross validated analysis in both. Quadratic discriminant analysis was used and rightfully found to be the best as our variables had different covariances. Quadratic discriminant analysis appears to be the most appropriate as it classified 0.98 of our data correctly.

### Question 3 

```{r question_3}
source("http://www.reuningscherer.net/multivariate/R/discrim.r.txt")
discriminant.significance(GED_Leveled[, c("Birth_Rate", "Completion_Rate_Upper_Secondary_Female")], GED_Leveled$ed_level)
```

* For 3 out of 4 of our variables the p-values from our Wilks test appear to be significant indicating that for these variables there is evidence of different mean literacy rates as predicted by our variables. The only variable that is not significant is Birth Rate which shows that birth rate is not a useful variable in discriminating for literacy. 

### Question 4

``` {r question_4}
source("http://www.reuningscherer.net/multivariate/R/discrim.r.txt")
discriminant.significance(GED_Leveled[,c("Completion_Rate_Upper_Secondary_Female", "Completion_Rate_Upper_Secondary_Male")], GED_Leveled$ed_level)
```
* We are detecting how well LDA discriminates between two groups. We initially see that completion rate upper secondary female and completion rate secondary male are significant—whereas birth rate and unemployment rate are not. Function 1 (Female copletion rate) appears to have better discriminating power than Function 2 (Male completion rate)because its Wilks' Lambda is lower (0.1790 compared to 0.9355). This makes sense because a country's efforts to raise female education levels will likely be more holistic for their broader population.

### Question 5
```{r}
GED.disc <- lda(GED_Leveled[, -5], grouping = GED_Leveled$ed_level)

(ctraw <- table(GED_Leveled$ed_level, predict(GED.disc)$class))
# total percent correct
round(sum(diag(prop.table(ctraw))),2)


# Quadratic Discriminant Analysis
GED.qdisc <- qda(GED_Leveled[, -5], grouping = GED_Leveled$ed_level)

(ctrawqed <- table(GED_Leveled$ed_level, predict(GED.qdisc)$class))

# total percent correct
round(sum(diag(prop.table(ctrawqed))),2)

# LDA cross validated results
GED.discCV <- lda(GED_Leveled[,-5], grouping = GED_Leveled$ed_level, CV = TRUE)
(ctCV <- table(GED_Leveled$ed_level, GED.discCV$class))
round(sum(diag(prop.table(ctCV))),2)

#Quadratic cross validated results
GED.discCVqed <- qda(GED_Leveled[,-5], grouping = GED_Leveled$ed_level, CV = TRUE)
(ctCVQ <- table(GED_Leveled$ed_level, GED.discCVqed$class))
round(sum(diag(prop.table(ctCVQ))),2)
```
* Analyzing the returns, it seems that our predictors were very strong. On regular and leave-on-out tests for both linear and quadratic discriminant analysis, we found strong prediction rates between .95 and .98. QDA, however, had slightly higher rates. This indicates to us that QDA predictors were especially strong, as we expected.

### Question 6
```{r}
(GED_Leveled_disc_lda)
```

* Upper secondary Female completion and upper secondary male completion has the most impact on educational level, due to having the largest absolute value. On the other hand, unemployment rate and birth rate did not seem to have strong impacts.

### Question 7
* As we utilize Quadratic Discriminant Analysis, scoreplots are not appropriate for our data.

### Question 8
```{r}
levels(GED_Leveled$ed_level) <- c("Low", "Medium", "High")
partimat(GED_Leveled$ed_level ~ Completion_Rate_Upper_Secondary_Female + Completion_Rate_Upper_Secondary_Male, data = GED_Leveled, method = "lda")
```
* Our graph here is spanned across two variables: completion rate of upper secondary school for females, and the completion rate of upper secondary school for males. It seems to be reasonably strong, with most of the lower levels of education (The light blue region) falling into one region only around 4 mistakes. The second region (the white region) seems to only have 5-6 mistakes. The third region (the pink region)  seems to only have 4-5 mistakes.

``` {r}

install.packages("class")
library(class)

```

### Question 9
```{r}

k <- 3

train_features <- GED_Leveled[, c("Birth_Rate","Completion_Rate_Upper_Secondary_Female", "Completion_Rate_Upper_Secondary_Male", "Unemployment_Rate")]

test_features <- GED_Leveled[, c("Birth_Rate","Completion_Rate_Upper_Secondary_Female", "Completion_Rate_Upper_Secondary_Male", "Unemployment_Rate")]

train_labels <- as.factor(GED_Leveled$ed_level)

true_labels <- as.factor(GED_Leveled$ed_level)

predicted_labels <- knn(train = train_features, test = test_features, cl = train_labels, k = k)

GED_Leveled$predicted_labels <- predicted_labels

ggplot(GED_Leveled, aes(x = Completion_Rate_Upper_Secondary_Female, y = Completion_Rate_Upper_Secondary_Male, color = predicted_labels, shape = ed_level)) +
  geom_point(size = 3) +
  labs(title = "Ed Level - KNN Classification Results", x = "Completion_Rate_Upper_Secondary_Female", y = "Completion_Rate_Upper_Secondary_Male", color = "Predicted Class", shape = "True Class") +
  scale_color_manual(values = c("red", "blue", "Orange")) + 
  scale_shape_manual(values = c(16, 17, 18)) + 
  theme_minimal()

```
